---
title: "simulation of the AIDS trial study"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(MASS)
library(wakefield)
library(survival)
 load("/Users/marcy/Downloads/SimulatedHIVPrevTrial_V2.Rdata")
```
In this R markdown file, we are trying to replicate the simulation posted in "Strategies for cohort sampling under the cox proportional hazards model, application to an AIDS clinical trial", by Kim and De Gruttola. A sample size of 1000 was selected to resemble ACTG 019, a phase III clinical trial. Here, we have two covariates, Z1 and Z2, that follows a bivariate normal distribution with mean equal to (0,0)' and both variances equals to 1, and correlation equals to 0.5. Failure times were generated according to an expoential distribution with hazard rho0(exp(Z1beta1+Z2beta2)). Some example covariates under this circustance could be: CD4 cell count and age. The first covariate Z1 is observed for the entire cohort, and Z2, in theory, is observed only for the failures and the selected controls. The cox regression parameters beta 1 and beta 2 were set to 1 and 0.5 respectively, and rho0 to 0.1. Times of entry into the study were generated by a uniform (0,1) distribution, hence time of censoring is also uniform (0,1). In the reference paper, the average number of failures was approximately 91, whereas in this simulation, the average number of failures was 91. For the case-cohort designs, sampling method was to obtain approximately one control per failure. No stratification or weighting method was used, and the parameter estimates were obtained by applying the Self and Prentice method.
```{r}
set.seed(1009)
# creating single function to get single failure count, pg 156 of the Kim and De Gruttola paper
failure.times <- function(n=1000,mu1=0,mu2=0,s1=1,s2=1,rho=0.5,beta1=1,beta2=0.5,rho0=0.1) {
#n <- 1000
##create bivariate normal distribution for Z1 and Z2, with mean=(0,0)', var=1 and cor = 0.5
mu1=mu2=0
s1=s2=1
rho <- 0.5
mu <- c(mu1,mu2) # Mean
sigma <- matrix(c(s1^2, s1*s2*rho, s1*s2*rho, s2^2),
           2,2) # Covariance matrix
bvn1 <- mvrnorm(n, mu,sigma, tol = 1e-6, empirical = FALSE, EISPACK = FALSE)
colnames(bvn1) <- c("Z1","Z2")
## generate failure times
beta1 <- 1
beta2 <- 0.5
rho0 <- 0.1
vals <- as.data.frame(bvn1)
rate <- rho0*(exp(vals$Z1*beta1+vals$Z2*beta2))
f.r <- rexp(1000, rate)
## time of entry into the study
t.e <- runif(n, min = 0, max = 1)
## time of censoring
t.c <- runif(n, min = 0, max = 1)
failures <- t.c > f.r
time.to.failure <- t.e+f.r
sum(failures, na.rm = TRUE)}
# testing the results
failure.times(n=1000,mu1=0,mu2=0,s1=1,s2=1,rho=0.5,beta1=1,beta2=0.5,rho0=0.1)
# use a wrapper for average failure times
ave.failure.times <- function(z){
  sep.counts <- replicate(z, failure.times(n=1000,mu1=0,mu2=0,s1=1,s2=1,rho=0.5,beta1=1,beta2=0.5,rho0=0.1)) 
  sum(sep.counts)/z
}
# testing the result, compare it with 91 (reference number)
f.times <- ave.failure.times(1000)
```
To use the cch package for estimating parameters. We obtained the estimated coefficient beta 1 to be 1.121, and the estimated coefficient beta 2 to be 0.453. 
```{r}
sim.data.generator <- function(n,mu1,mu2,s1,s2,rho,beta1,beta2,rho0) {
id <- seq(1,n,1)
n
mu1=mu2
s1=s2
mu <- c(mu1,mu2) # Mean
sigma <- matrix(c(s1^2, s1*s2*rho, s1*s2*rho, s2^2),
           2,2) # Covariance matrix
bvn1 <- mvrnorm(n, mu,sigma, tol = 1e-6, empirical = FALSE, EISPACK = FALSE)
colnames(bvn1) <- c("Z1","Z2")
## generate failure times
vals <- as.data.frame(bvn1)
rate <- rho0*(exp(vals$Z1*beta1+vals$Z2*beta2))
failure.times <- rexp(n, rate)
## time of entry into the study
time.of.entry <- runif(n, min = 0, max = 1)
## time of censoring
time.of.censoring <- runif(n, min = 0, max = 1)
failures <- time.of.censoring > failure.times
time.to.failure <- time.of.entry+failure.times
data.frame(id, bvn1, failure.times, time.of.entry, time.of.censoring, failures, time.to.failure)
}
sim.data <- sim.data.generator(n=1000,mu1=0,mu2=0,s1=1,s2=1,rho=0.5,beta1=1,beta2=0.5,rho0=0.1)

# sample from the full study cohort to obtain approximately one control per failure
# random.samp <- sample(n, f.times, replace = FALSE)
# randomly sample 92 TRUES (in subcohort)
a <- 182
cch.w.samp <- function(a,b){
in.sub <- sample(c(rep(TRUE, a), rep(FALSE, b-a)), b ,replace = F)
# forming the simulation dataset
sim.sub <- data.frame(sim.data, in.sub)
# subletting the in.sub variable (indicator of sampling into the subcohort)
subcoh <- sim.sub$in.sub
selccoh <- with(sim.sub, failures==1|in.sub==1) ## cases in the subcohort
# creating the subcohort dataset from the full cohort dataset
sim.subco <- sim.sub[selccoh,]
# run subcohort analysis using the cch function with SP method
fit1 <- cch(Surv(time.to.failure, failures) ~ Z1+Z2, data =sim.subco,
   subcoh = ~in.sub, id=~id, cohort.size=1000, method="SelfPren")
means.beta1 <- fit1$coefficients[1]
vars.beta1 <- fit1$var[1]
means.beta2 <- fit1$coefficients[2]
vars.beta2 <- fit1$var[4]
list(means.beta1,vars.beta1,means.beta2,vars.beta2)
}
res.list <- as.numeric(replicate(1000, cch.w.samp(182,1000)))
B = matrix(res.list, nrow=1000, ncol=4,byrow=TRUE) 
est.beta1.means <- as.numeric(B[,1])
est.beta1.vars <- as.numeric(B[,2])
est.beta2.means <- as.numeric(B[,3])
est.beta2.vars <- as.numeric(B[,4])
beta1.avg.mean <- mean.default(est.beta1.means)
beta1.avg.var <- mean.default(est.beta1.vars)
beta1.samp.var <- var(est.beta1.means)
beta2.avg.mean <- mean.default(est.beta2.means)
beta2.avg.var <- mean.default(est.beta2.vars)
beta2.samp.var <- var(est.beta2.means)


res.matrix <- matrix(c(beta1.avg.mean,beta1.avg.var,beta1.samp.var,
                       beta2.avg.mean,beta2.avg.var,beta2.samp.var,
                       1.18, 0.084, 0.180,0.58, 0.069, 0.123),ncol=6,byrow=T)
colnames(res.matrix) <- c("beta1.avg.mean","beta1.avg.var","beta1.samp.var",
                          "beta2.avg.mean","beta2.avg.var","beta2.samp.var")
rownames(res.matrix) <- c("obtained values","listed values")
res.table <- as.table(res.matrix)
```
Next, we used the code from the MORGAM project paper, without the attach and detach commends, and set the probability to be 1 for all participants. The estimation for beta 1 is 1.1, and the estimation for beta 2 if 0.4. The number of failures obtained from these commends is 86.
```{r}
ccregression <- function(dataset,covariates,  idvar, censvar, agestart, agestop, prob, subcoh) {
   n <- nrow(dataset)
   epsilon <- 0.00001
    start <- NULL

   stop <- NULL

   cens <- NULL

   weight <- NULL

   keys <- NULL
   

z <- matrix(NA, n, length(covariates)) # the covariate matrix

for (i in 1:length(covariates))
  ## subset the dataset, set the col names to be each element of covariates
  z[,i] <- as.numeric(dataset[,names(dataset) == covariates[i]]) 
## set the column names
colnames(z) <- covariates

   for (i in 1:n) {

      # Case outside subcohort
          if ((censvar[i]) & (!subcoh[i])) { ##if inside case and not variable in the subcohort

            start <- c(start, agestop[i]-epsilon)

            stop <- c(stop, agestop[i])

            cens <- c(cens, 1)

            weight <- c(weight, 1)

            keys <- c(keys, idvar[i]) 

      }

      # Non-case in subcohort

       else if ((!censvar[i]) & (subcoh[i])) {

            start <- c(start, agestart[i])

            stop <- c(stop, agestop[i])

            cens <- c(cens, 0)

            weight <- c(weight, 1/prob[i])

            keys <- c(keys, idvar[i])

      }

      # Case in subcohort

      else if ((censvar[i]) & (subcoh[i])) {

            start <- c(start, agestart[i])

            stop <- c(stop, agestop[i]-epsilon)

            cens <- c(cens, 0)

            weight <- c(weight, 1/prob[i])

            keys <- c(keys, idvar[i])

            start <- c(start, agestop[i]-epsilon)

            stop <- c(stop, agestop[i])

            cens <- c(cens, 1)

            weight <- c(weight, 1)

            keys <- c(keys, idvar[i])

         }

      }

      y <- Surv(start, stop, cens) ##time is start, time2 is stop, event is cens (case status)

     z_ <- z[match(keys, idvar),]

      return(coxph(y ~ z_+ cluster(as.factor(keys)), weights=weight))
cox.fit <- coxph(y ~ z_+ cluster(as.factor(keys)), weights=weight)
summary(cox.fit)
}
covariates1 <- c("Z1","Z2")
ccregression(sim.sub, covariates1, idvar, failures, t.e, time.to.failure, prob, in.sub)

```
To practice with the code from the MORGAM project, we generated another simulation dataset that mimics the MORGAM project, and try to use both cch package and the modified MORGAM code to fit a model and obtain parameter estimates. The dataset has 6 variables: ptid is the patient id variable; age is the covariate in this dataset; site is a categorical variable that could be used in stratification; enroll.day is the start of the study; last.fu.day is the end of the study; hiv.infection is the outcome of interest (event). We want to graph the realtionship between sub.cohort size and variances using the SP method, for cch and MORGAM code.
```{r}
## try subcohort size of 500, random sampling
n2 <- 2500
sel.sub <- sample(c(rep(TRUE, 500), rep(FALSE, n2-500)), n2 ,replace = F)
sim.trial$sel.sub <- sel.sub
st.subcoh <- sim.trial$sel.sub
st.selccoh <- with(sim.trial, hiv.infection==1|sel.sub==1) ## cases in the subcohort
# creating the subcohort dataset from the full cohort dataset
st.sim.subco <- sim.trial[st.selccoh,]
# run subcohort analysis using the cch function with SP method
st.fit <- cch(Surv(last.fu.day, hiv.infection) ~ age+int, data =st.sim.subco,
   subcoh = ~sel.sub, id=~ptid, cohort.size=2500, method="SelfPren")
summary(st.fit)
```

```{r}
## altering the subcohort size, plot variance vs. the subcohort size
cchfcn <- function(n) {
  sel.sub <- sample(c(rep(TRUE, n), rep(FALSE, n2-n)), n2 ,replace = F)
  sim.trial$sel.sub <- sel.sub
  st.subcoh <- sim.trial$sel.sub
  st.selccoh <- with(sim.trial, hiv.infection==1|sel.sub==1)
  st.sim.subco <- sim.trial[st.selccoh,]
  st.fit <- cch(Surv(last.fu.day, hiv.infection) ~ age+site, data =st.sim.subco,
   subcoh = ~sel.sub, id=~ptid, cohort.size=2500, method="SelfPren")
  st.fit$var[1]
}
cchfcn(500)
## to examine the relationship for subcohort size of 30 to 2500 (full cohort)
nset <- seq(30, 2500, 1)
varset <- lapply(nset, cchfcn)
p0 <- plot(nset, varset, xlab= "subcohort size (30-2500)", ylab = "variances", main="Variances vs. subcohort size")
## to limit the range to 100~1000
nset.2 <- seq(100, 1000, 1)
varset.2 <- lapply(nset.2, cchfcn)
p <- plot(nset.2, varset.2, xlab= "subcohort size (30-1000)", ylab = "variances", main="Variances vs. subcohort size")
## to get the relative efficiency: variances of the full cohort / variances of the subcohort cohort 
full.fit <- coxph(Surv(last.fu.day, hiv.infection)~age+site, data=sim.trial)
attributes(full.fit)
full.fit$var[1] ## obtaining the variance of the full cohort
cchfcn.ef <- function(n) { ##to get the efficiency
  sel.sub <- sample(c(rep(TRUE, n), rep(FALSE, n2-n)), n2 ,replace = F)
  sim.trial$sel.sub <- sel.sub
  st.subcoh <- sim.trial$sel.sub
  st.selccoh <- with(sim.trial, hiv.infection==1|sel.sub==1)
  st.sim.subco <- sim.trial[st.selccoh,]
  st.fit <- cch(Surv(last.fu.day, hiv.infection) ~ age+site, data =st.sim.subco,
   subcoh = ~sel.sub, id=~ptid, cohort.size=2500, method="SelfPren")
  full.fit$var[1]/st.fit$var[1]
}
cchfcn.ef(500)
## to examine the relationship between relative efficiency and subcohort sizefor subcohort size of 30 to 2500 (full cohort)
nset.ef <- seq(30, 2500, 1)
varset.ef <- lapply(nset.ef, cchfcn.ef)
p1 <- plot(nset.ef, varset.ef, xlab= "subcohort size (30-2500)", ylab = "relative efficiency", main="Relative efficiency vs. subcohort size")
## to limit the range to 30~500
nset.2.ef <- seq(30, 500, 1)
varset.2.ef <- lapply(nset.2.ef, cchfcn.ef)
p2 <- plot(nset.2.ef, varset.2.ef)
```

```{r}
prob <- rep(1,2500)
sim.trial$prob <- prob
st.covariates <- c("age","int")
ccregression(sim.trial, st.covariates, sim.trial$ptid, sim.trial$hiv.infection, sim.trial$enroll.day, sim.trial$last.fu.day, sim.trial$prob, sim.trial$sel.sub)




```
Practicing sampling based on age distribution
According to the MORGAM paper, the number of major endpoint events during the follow-up increases strongly with the age of the individual at baseline. So the analysis will not be powered if all members of the cohort had an equal probability of being selected, since the average age of the cases would be much higher than the average age of the subcohort. By adjusting the selecting probability based on age, the power can be increase to a level comparable to the power of an age-matched case-control design by selecting the individuals of the subsample using age-distribution similar to the distribution of the baseline age of the cases. In the paper, such a function of age is the mortality rate estimated using a logistic regression model. In our case, we are using hiv.infection as the outcome variable.
```{r}
library(uwIntroStats)
regress("odds", hiv.infection~age+factor(site), data=sim.trial)
## from the logistic regression, we get alpha = -2.237, beta = -3.795e-03
## The paper pointed out that an individual with age bi at baseline is selected for the sample with probability proportional to 
## f(bi) = exp(alpha+beta*bi)/(1+exp(alpha+beta*bi))
## following pg 5 of the MORGAM paper
prob.ind <- function(age) {
  alpha = -2.237
  beta = -3.795e-03
  f = exp(alpha+beta*age)/(1+exp(alpha+beta*age))
  f
}
prob.ind(30)
results <- rep(NA,2500)
p.sel <- rep(NA,2500)
 for (i in 1:2500) {
   results[i] <- prob.ind(sim.trial$age[i])
 }
## let 600 be the subcohort size
full.co.sum <- sum(results)
for (i in 1:2500){
   p.sel[i] <- 250*results[i]/full.co.sum
}

sim.trial$age.prob <- p.sel
plot(sim.trial$age, p.sel)
age.sel.subco <- rep(NA,2500)
for (i in 1:2500) {
  age.sel.subco[i] <- rbinom(1,1, sim.trial$age.prob[i])
}
sim.trial$age.sel.subco <- age.sel.subco
```
MOGRAM code
```{r}
## age distribution adjusted weight
covariates1 <- c("age","int")
ccregression(sim.trial, covariates1, sim.trial$ptid, sim.trial$hiv.infection, sim.trial$enroll.day, sim.trial$last.fu.day, sim.trial$age.prob, sim.trial$age.sel.subco)
```
CCH package
```{r}
age.subcoh <- sim.trial$age.sel.subco
age.selccoh <- with(sim.trial, hiv.infection==1|age.sel.subco==1) ## cases in the subcohort
# creating the subcohort dataset from the full cohort dataset
age.sim.subco <- sim.trial[age.selccoh,]
# run subcohort analysis using the cch function with SP method
age.fit <- cch(Surv(last.fu.day, hiv.infection) ~ age+int, data =age.sim.subco,
   subcoh = ~age.sel.subco, id=~ptid, cohort.size=2500, method="SelfPren")
summary(age.fit)

```

```{r}
plot(density(sim.trial$age[sim.trial$age.sel.subco==1]))
lines(density(sim.trial$age[sim.trial$hiv.infection==1]))
lines(density(sim.trial$age))
par(mfrow=c(1,2))
plot(nset, varset, xlab= "subcohort size (30-2500)", ylab = "variances", main="Variances vs. subcohort size")
plot(nset.ef, varset.ef, xlab= "subcohort size (30-2500)", ylab = "relative efficiency", main="Relative efficiency vs. subcohort size")
```